---
title: "Thesis"
author: "Jared Keirn"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
# set.seed(123) # Uncomment for reproducibility
```

# Introduction

This notebook explores the prevalence of Adverse Childhood Experiences (ACEs) among adolescent and young women in Kenya, and associations with adulthood risk behavior.

# Libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(skimr)  
library(naniar) 
```

# Data Import

```{r load-data}
# Use a relative path for reproducibility
data_path <- "/Users/jaredkeirn/Dropbox/My Mac (Jrod)/Desktop/Thesis/AGYWPharmacyPrEPRCT-JKExport_DATA_2025-05-28_1825.csv"
agyw_redcap <- read_csv(data_path)

# Quick data overview
head(agyw_redcap)
summary(agyw_redcap)
str(agyw_redcap)
```

# Event Order (Chronological)

```{r event-order}
event_order <- c(
  "enrollment_arm_1",
  "1_month_fup_arm_1",
  "4_month_fup_arm_1",
  "7_month_fup_arm_1",
  "10_month_fup_arm_1",
  "interim_visit_arm_1",
  "termination_arm_1"
)
```

# Collapse to One Row Per Participant

```{r collapse-data}
prioritize_enrollment <- function(x, event) {
  enroll_val <- x[event == "enrollment_arm_1"]
  if (length(enroll_val) > 0 && !is.na(enroll_val[1])) {
    return(enroll_val[1])
  } else {
    return(na.omit(x)[1])
  }
}

agyw_collapsed <- agyw_redcap %>%
  group_by(clt_ptid) %>%
  summarise(
    across(starts_with("ace_"), ~ na.omit(.)[1], .names = "{.col}"),
    across(
      .cols = c(-redcap_event_name, -starts_with("ace_")),
      .fns = ~ prioritize_enrollment(., redcap_event_name), 
      .names = "{.col}"
    ),
    .groups = 'drop'
  )

head(agyw_collapsed)
summary(agyw_collapsed)
```

# Filter Participants (Aged 18+) and Drop REDCap Admin Columns

```{r filter-clean}
agyw_filtered <- agyw_collapsed %>%
  filter(!is.na(dem_age) & dem_age >= 18) %>%
  select(-starts_with("redcap_"))
head(agyw_filtered)
summary(agyw_filtered)
```

# Explore ACE Variable Frequencies

```{r ace-frequencies}
ace_columns <- names(agyw_filtered)[startsWith(names(agyw_filtered), "ace_")]

for (col in ace_columns) {
  print(paste("Frequency for column:", col))
  kable(as.data.frame(table(agyw_filtered[[col]], useNA = "ifany")),
        caption = col)
}
```

# Recode Special Codes (`77 = Refused`, `88 = Not sure`) as NA

```{r recode-special-codes}
knitr::kable(data.frame(Column = names(agyw_filtered)), caption = "List of Columns")

col_with_77 <- c("ace_married", "ace_marchoose", "ace_marconsent", "ace_parprob",
                 "ace_parknow","ace_parfood","ace_pardrunk", "ace_parschool",
                 "ace_famaddict","ace_famdepr","ace_famjail", "ace_pardivor",
                 "ace_pardie", "ace_hminsult","ace_attemptintercorse","ace_abandon",
                 "ace_hmviolence", "ace_forceintercourse", "ace_hmviolenceobj", "ace_insult",
                 "ace_violence", "ace_bullied", "ace_violenceobj", "ace_bullymeth", "ace_fight",
                 "ace_sexassult", "ace_cmbeat", "ace_forcetouch", "ace_cmstab", "ace_cmthreat", "ace_move",
                 "ace_destruct", "ace_colbeat","ace_colfmviol", )

col_with_88 <- c("ace_marchoose", "ace_pardivor", "ace_pardie")

agyw_filtered <- agyw_filtered %>%
  mutate(
    across(all_of(col_with_77), ~na_if(., 77)),
    across(all_of(col_with_88), ~na_if(., 88))
  )
```

# Rename Age Variable

```{r rename-age}
agyw_trans <- agyw_filtered %>%
  rename(Age_years = dem_age)
```

# Recoding Example (Bundle Multiple Recodes)

```{r ace-recoding}
agyw_trans <- agyw_trans %>%
  mutate(
    ace_parprob_yn = ifelse(ace_parprob %in% c(1:4), 0, ifelse(ace_parprob == 5, 1, NA)),
    ace_parknow_yn = case_when(
      ace_parknow %in% c(1:4) ~ 0,
      ace_parknow == 5 ~ 1,
      TRUE ~ NA_real_
    ),
    ace_parfood_yn = case_when(
      ace_parfood == 1 ~ 1,
      ace_parfood %in% c(2,3,0) ~ 0,
      TRUE ~ NA_real_
    ),
    ace_pardrunk_yn = case_when(
      ace_pardrunk == 1 ~ 1,
      ace_pardrunk %in% c(2,3,0) ~ 0,
      TRUE ~ NA_real_
    )
    # ...continue for other recodes...
  )
```

# Inspect and Summarize Key Variables

```{r inspect-key-vars}
kable(table(agyw_trans$ace_married, useNA = "ifany"), caption = "Ever Married?")
kable(table(agyw_trans$ace_marriedage, useNA = "ifany"), caption = "Age First Married")
kable(table(agyw_trans$ace_parprob_yn, useNA = "ifany"), caption = "Parental Understanding (YN)")
kable(table(agyw_trans$ace_parknow_yn, useNA = "ifany"), caption = "Parental Knowledge (YN)")
```

# Covariates and Outcomes (Descriptives)

```{r covariates-outcomes}
kable(table(agyw_trans$dem_school_year, useNA = "ifany"), caption = "Years of Education")
kable(table(agyw_trans$dem_area_live_in, useNA = "ifany"), caption = "Area Type")
kable(table(agyw_trans$dem_formally_employed, useNA = "ifany"), caption = "Formally Employed")
kable(table(agyw_trans$dem_without_condom, useNA = "ifany"), caption = "Had Sex Without Condom")
kable(table(agyw_trans$dem_sex_money, useNA = "ifany"), caption = "Sex in Exchange for Money/Favors")
kable(table(agyw_trans$dem_intravenous_drugs, useNA = "ifany"), caption = "Intravenous Drug Use")
```

# Save Processed Data

```{r save-processed}
saveRDS(agyw_trans, file = "data/agyw_processed.rds")
print("Data frame saved successfully to: data/agyw_processed.rds")
```

# (Optional) Visualize Missing Data

```{r missing-data-visual, eval=FALSE}
# library(naniar)
# gg_miss_var(agyw_trans, show_pct = TRUE)
```

# (Optional) Skim Summary

```{r skim-summary, eval=FALSE}
# library(skimr)
# skim(agyw_trans)
```

# Session Info

```{r session-info}
sessionInfo()
```

# Conclusion

This notebook processes, recodes, and summarizes the AGYW REDCap data for further analysis of ACEs and risk behavior.\
For further modeling or visualization, add new sections below.
